\vspace{-1cm} \normalsize Supplementary Information\\\vspace{0.2cm} \LARGE Basic aspects of species' distributions

### 2021-06-23 manuscript/manuscript.tex
New simulations

To test the performance of the model as well as our choice of prior distributions, we modelled simulated data and compared the sampled posterior distributions to the data-generating parameters (see Code Availability section). Notice that using the link function in Eq.~(\ref{eq:baseline-response}) could cause problems when sampling the model, and some adjustments need to be made when specifying the model (see Code Availability section). To perform the data analysis and generate the figures, we used some of the functions available with the R package `rethinking' \citep{mcelreathStatisticalRethinkingBayesian2020}.

@@

\section*{Results}
We studied the distribution data to characterize species' realized niches along the main axis of variation of all environmental variables. Using the presence and absence of species across sites as the response variable, we sampled the posterior distributions of the baseline model, accounting for the information in the floristic database regarding species' indicator values and range of variation. This allowed us to map the center and variance of species' distributions along the environmental gradient (Fig.~\ref{fig:correlation}).  Studying the relationship between these properties, we found these to be negatively correlated (i.e.~$\beta_{i}$ and $\gamma_{i}$ in the baseline model were positively correlated; Fig.~\ref{fig:correlation}). This means that species found at lower elevations have generally wider distributions than those at higher elevations. The same relationship was found when using instead elevation or mean temperature as explanatory variables (Supplementary Fig.~3) as well as when using ordinal data (Supplementary Fig.~4); however, the pattern was not present along the second main axis of variation of our environmental variables (Supplementary Fig.~5). The comparison between the other parameter estimates revealed additional, somewhat more expected, relationships. In particular, we found the amplitude of distributions to be positively and negatively correlated with their mean and variance, respectively (i.e. $\alpha_{i}$ is positively correlated with $\beta_{i}$ and $\gamma_{i}$; Supplementary Fig.~XX). This implies that, at higher elevations, species' distributions generally have lower amplitudes.

@@

Maintaining the symmetry of species' distributions, we then allowed the kurtosis---or shape of the tails---of these to vary in different ways. To do so, we changed the response curve of our Bayesian model to follow a generalized error distribution (Eq.~\ref{eq:generalized-response}). A comparison of the WAIC values showed this non-linear regression to outperform the baseline model (Supplementary Fig.~XX). Studying the resulting posterior distributions, we found the average kurtosis of the distributions to be slightly greater than zero, which corresponds to distributions with longer tails than the Gaussian case (Fig.~\ref{fig:kurtosis-skewness}). However, the parameter controlling for the kurtosis $\nu_i$ displayed a lot of variation across species (Supplementary Fig.~XX), which might indicate that the shape of the tails is species-specific.

\begin{figure}[ht]
  \centering
    \includegraphics[width=0.9\textwidth]{figures/kurto-skew}
    	  \vspace{0.1cm}
	   \caption{Average kurtosis and skewness of species' distributions. Calculated using the posterior distributions of parameters $\hat{\nu}$ and $\hat{\lambda}$ from the models (see Supplementary Table XX), the two panels describe the average (a) kurtosis and (b) skewness of distributions. Panel (a) displays the results obtained using response curves that follow a generalized error distribution (black line) and a skewed generalized error distribution (yellow line). Panel (b) displays the results obtained using response curves that follow a skewed normal distribution (black line) and a skewed generalized error distribution (yellow line). In both cases, the gray dotted line indicates the conditions by which species are normally distributed along the environmental axis.}
      \label{fig:kurtosis-skewness}
      %7.5-3
\end{figure}

Using Eq.~(\ref{eq:skewed-response}), we next studied the skewness of species' distributions. Based on the estimates for the WAIC values, this model outperformed the first two (Supplementary Fig.~XX), which sheds light on the naturally skewed nature of species' distributions. Perhaps most importantly, studying the mean value of the skewness across species, we found this to be consistently below zero (Fig.~\ref{fig:kurtosis-skewness}). This indicates that species' distributions generally present steeper declines towards higher elevations (i.e. $\hat{\lambda}<0$; Fig.~\ref{fig:response}). The same was true when using a model that allowed for both fat-tailed and skewed response curves (Eq.~\ref{eq:heavy-skewed-response}). This model outperformed the rest, presenting Akaike weights close to 1 (Supplementary Fig.~XX), suggesting that both the kurtosis and skewness are useful properties to describe empirical distributions (Fig.~\ref{fig:kurtosis-skewness}). A study of how the prior knowledge we had regarding species' environmental preferences, range of variation and ecological strategies informed the different parameters of the model is presented in the Supplementary Note XX. %Comparing the prior and posterior distributions for the hyperparameters in the.

>>>>

To test the performance of the model as well as our choice of prior distributions, we modelled simulated data and compared the sampled posterior distributions to the data-generating parameters (e.g.~Supplementary Fig.~2; see Code Availability section). Notice that using the link function in Eq.~(\ref{eq:baseline-response}) could cause problems when sampling the model, and some adjustments need to be made when specifying the model (see Code Availability section). To perform the data analysis and generate the figures, we used some of the functions available with the R package `rethinking' \citep{mcelreathStatisticalRethinkingBayesian2020}.

@@

\section*{Results}
We studied the distribution data to characterize species' realized niches along the main axis of variation of all environmental variables. Using the presence and absence of species across sites as the response variable, we sampled the posterior distributions of the baseline model, accounting for the information in the floristic database regarding species' indicator values and range of variation. This allowed us to map the center and variance of species' distributions along the environmental gradient (Fig.~\ref{fig:correlation}).  Studying the relationship between these properties, we found these to be negatively correlated (i.e.~$\beta_{i}$ and $\gamma_{i}$ in the baseline model were positively correlated; Fig.~\ref{fig:correlation}). This means that species found at lower elevations have generally wider distributions than those at higher elevations. The same relationship was found when using instead elevation or mean temperature as explanatory variables (Supplementary Fig.~3) as well as when using ordinal data (Supplementary Fig.~4); however, the pattern was not present along the second main axis of variation of our environmental variables (Supplementary Fig.~5). The comparison between the other parameter estimates revealed additional, somewhat more expected, relationships. In particular, we found the amplitude of distributions to be positively and negatively correlated with their mean and variance, respectively (i.e. $\alpha_{i}$ is positively correlated with $\beta_{i}$ and $\gamma_{i}$; Supplementary Fig.~6). This implies that, at higher elevations, species' distributions generally have lower amplitudes.

@@

Maintaining the symmetry of species' distributions, we then allowed the kurtosis---or shape of the tails---of these to vary in different ways. To do so, we changed the response curve of our Bayesian model to follow a generalized error distribution (Eq.~\ref{eq:generalized-response}). A comparison of the WAIC values showed this non-linear regression to outperform the baseline model (Supplementary Fig.~7). Studying the resulting posterior distributions, we found the average kurtosis of the distributions to be slightly greater than zero, which corresponds to distributions with longer tails than the Gaussian case (Fig.~\ref{fig:kurtosis-skewness}). However, the parameter controlling for the kurtosis $\nu_i$ displayed a lot of variation across species (Supplementary Fig.~8), which might indicate that the shape of the tails is species-specific.

\begin{figure}[ht]
  \centering
    \includegraphics[width=0.9\textwidth]{figures/kurto-skew}
    	  \vspace{0.1cm}
	   \caption{Average kurtosis and skewness of species' distributions. Calculated using the posterior distributions of parameters $\hat{\nu}$ and $\hat{\lambda}$ from the models (see Supplementary Table 2 and \citealt{vignette}), the two panels describe the average (a) kurtosis and (b) skewness of distributions. Panel (a) displays the results obtained using response curves that follow a generalized error distribution (black line) and a skewed generalized error distribution (yellow line). Panel (b) displays the results obtained using response curves that follow a skewed normal distribution (black line) and a skewed generalized error distribution (yellow line). In both cases, the gray dotted line indicates the conditions by which species are normally distributed along the environmental axis.}
      \label{fig:kurtosis-skewness}
      %7.5-3
\end{figure}

Using Eq.~(\ref{eq:skewed-response}), we next studied the skewness of species' distributions. Based on the estimates for the WAIC values, this model outperformed the first two (Supplementary Fig.~7), which sheds light on the naturally skewed nature of species' distributions. Perhaps most importantly, studying the mean value of the skewness across species, we found this to be consistently below zero (Fig.~\ref{fig:kurtosis-skewness}). This indicates that species' distributions generally present steeper declines towards higher elevations (i.e. $\hat{\lambda}<0$; Fig.~\ref{fig:response}). The same was true when using a model that allowed for both fat-tailed and skewed response curves (Eq.~\ref{eq:heavy-skewed-response}). This model outperformed the rest, presenting Akaike weights close to 1 (Supplementary Fig.~7), suggesting that both the kurtosis and skewness are useful properties to describe empirical distributions (Fig.~\ref{fig:kurtosis-skewness}). A study of how the prior knowledge we had regarding species' environmental preferences, range of variation and ecological strategies informed the different parameters of the model is presented in the Supplementary Note 1 and Supplementary Fig.~9. %Comparing the prior and posterior distributions for the hyperparameters in the.

### 2021-06-23 manuscript/supplementary_information.tex
baseline model 2d
New simulations

\begin{figure}[ht]
  \centering
    \vspace{0.5cm}
    \includegraphics[width=1\textwidth]{figures/prior}
    	  \vspace{0.3cm}
	   \caption{Study of the priors for the baseline model presented in the main text.}
      \label{sfig:sensitivity}
\end{figure}

@@

\begin{figure}[h]
  \centering
    \vspace{0.5cm}
    \includegraphics[width=1\textwidth]{figures/figure1-secondaxis}
    	  \vspace{0.3cm}
	   \caption{Relationship between mean and variance of species' distributions. These are the results for the second axis of variation for the climatic data.}
      \label{sfig:sensitivity}
\end{figure}

\begin{figure}[ht]
  \centering
    \includegraphics[width=0.9\textwidth]{figures/means-Rapopor}
    	  \vspace{0.1cm}
	   \caption{Universality of the relationship between mean and variance of species' distributions. Comparison between how different types of species are mapped in Fig.~\ref{fig:correlation}a. Panel (a) describes the correlation coefficient between $\beta_i$ and $\gamma_i$ for each type of species. Panel (b) shows the differences between neophytes and native species in the way these are distributed along the environmental gradient. Panel (c) shows the same differences for species that have decreased, decreased in low elevations, increase and remain stable over the last decades (see Supplementary Table 1 for further details).}
	   %7.5-4
      \label{sfig:universality}
\end{figure}

>>>>

%\begin{figure}[ht]
%  \centering
%    \vspace{0.5cm}
%    \includegraphics[width=1\textwidth]{figures/prior}
%    	  \vspace{0.3cm}
%	   \caption{Study of the priors for the baseline model presented in the main text.}
%     \label{sfig:sensitivity}
%\end{figure}
%\clearpage

\begin{figure}[ht]
  \centering
    \vspace{0.5cm}
    \includegraphics[width=1\textwidth]{figures/simulations-comparison}
    	  \vspace{0.3cm}
	   \caption{Studying the simulations. Transparency is a way to indicate the different sample sizes. Say that for both $\beta$ and $gamma$, the posterior distributions for the different species are correlated so that the variance between species increases as they are further apart.}
      \label{sfig:simulations}
\end{figure}

@@

\begin{figure}[h]
  \centering
    \includegraphics[width=0.9\textwidth]{figures/categorical-figure1}
    	  \vspace{0.1cm}
	   \caption{Universality of the relationship between mean and variance of species' distributions. Comparison between how different types of species are mapped in Fig.~\ref{fig:correlation}a. Panel (a) describes the correlation coefficient between $\beta_i$ and $\gamma_i$ for each type of species. Panel (b) shows the differences between neophytes and native species in the way these are distributed along the environmental gradient. Panel (c) shows the same differences for species that have decreased, decreased in low elevations, increase and remain stable over the last decades (see Supplementary Table 1 for further details).}
	   %7.5-4
      \label{sfig:categorical-baseline}
\end{figure}

\begin{figure}[h]
  \centering
    \vspace{0.5cm}
    \includegraphics[width=1\textwidth]{figures/figure1-secondaxis}
    	  \vspace{0.3cm}
	   \caption{Relationship between mean and variance of species' distributions. These are the results for the second axis of variation for the climatic data.}
      \label{sfig:secondaxis}
\end{figure}

\begin{figure}[h]
  \centering
    \vspace{0.5cm}
    \includegraphics[width=0.9\textwidth]{figures/alpha-vs-gamma-beta}
    	  \vspace{0.3cm}
	   \caption{Relationship between mean and variance of species' distributions. These are the results for the second axis of variation for the climatic data.}
      \label{sfig:alpha-vs-gammabeta}
\end{figure}

\begin{figure}[h]
  \centering
    \vspace{0.5cm}
    \includegraphics[width=0.8\textwidth]{figures/WAIC-plot}
    	  \vspace{0.3cm}
	   \caption{Comparison of the different models using WAIC values. Each row describes one of the models described in the main text. (a) WAIC comparison of the different models. The filled points are the in-sample deviance of each model, and the open points are the estimated WAIC values. The dark line segments represent the standard error of each WAIC. The standard error of the difference between each WAIC and the best performing model (top row) is characterized by the gray triangles and line segments. The vertical line references the model with the lowest WAIC value. (b) Akaike weight for each model used to compare predictive accuracy of the different Bayesian models (see \citealt{mcelreathStatisticalRethinkingBayesian2020}). According to this measure, the model with fat-tails and skewed response is the best model. The open points are the mean Akaike weights and the line segments represent the corresponding standard error. The vertical line references the model with the highest Akaike weight. These graphs were generated using some of the functions from the R package \textit{rethinking} \citep{mcelreathStatisticalRethinkingBayesian2020}.}
      \label{sfig:WAIC}
\end{figure}

\begin{figure}[h]
  \centering
    \vspace{0.5cm}
    \includegraphics[width=0.8\textwidth]{figures/nu-kurtosis}
    	  \vspace{0.3cm}
	   \caption{Variability in the tails.}
      \label{sfig:nu-kurtosis}
\end{figure}
